---
title: "TODO - a title that states the finding"
description: "TODO - one line"
author: "Yipin Zhang"
date: 2026-09-21
categories: [web scraping, rvest, labor market]
execute: {warning: false, message: false}
---

```{r}
#| label: setup
#| include: false

library(dplyr)
library(ggplot2)
library(readr)
library(knitr)
library(here)

# Reads saved files only. The scrape lives in R/01_scrape.R and is run
# by hand, so rendering this post never contacts Hacker News.
POST_DIR <- here("posts", "web-scraping")

postings <- read_csv(file.path(POST_DIR, "data", "processed", "postings.csv"),
                     show_col_types = FALSE)
trend    <- read_csv(file.path(POST_DIR, "results", "tables", "skill_trends.csv"),
                     show_col_types = FALSE)

n_2025   <- sum(postings$thread == "2025-09")
n_2026   <- sum(postings$thread == "2026-09")
n_sig    <- sum(trend$p_value < 0.05, na.rm = TRUE)
r_count  <- trend$n_2026[trend$skill == "R"]
```

## The question

WRITE HERE. ~100 words. You write R for every assignment; R appears in
`r r_count` of `r n_2026` postings. What do employers actually ask for,
and has it shifted enough in a year to change what you should study?

## Where the data comes from

WRITE HERE. ~120 words. HN monthly hiring thread, one top-level comment
per employer. Same month a year apart holds seasonality fixed. robots.txt
allows /item, disallows /submitted, 30-second delay, pages cached. One
sentence on the API.

## Turning comments into data

WRITE HERE. ~120 words. 863 comments to 560 postings. The word-boundary
problem: \bsql\b misses PostgreSQL; a case-insensitive \bGo\b matched
the verb "go" until it was caught.

```{r}
#| label: fig-demand
#| fig-cap: "Most-requested technologies, September 2026"

trend |>
  slice_max(share_2026, n = 12) |>
  ggplot(aes(100 * share_2026, reorder(skill, share_2026))) +
  geom_col(fill = "steelblue") +
  labs(x = "% of postings mentioning the skill", y = NULL,
       subtitle = paste0("Share of ", n_2026, " job postings")) +
  theme_minimal(base_size = 12)
```

## What the comparison can and cannot show

WRITE HERE. ~180 words. TypeScript and React fell together, the two
largest moves. But `r n_sig` of the changes reach p < 0.05. Levels are
measured well; changes stack two standard errors.

```{r}
#| label: fig-change
#| fig-cap: "Change in share of postings, September 2025 to September 2026. No change is distinguishable from sampling noise."

trend |>
  filter(n_2025 + n_2026 >= 20) |>
  mutate(significant = p_value < 0.05) |>
  ggplot(aes(change_pp, reorder(skill, change_pp), fill = significant)) +
  geom_col() +
  geom_vline(xintercept = 0, colour = "grey30") +
  scale_fill_manual(values = c("TRUE" = "steelblue", "FALSE" = "grey78"),
                    labels = c("TRUE" = "p < 0.05", "FALSE" = "within noise"),
                    name = NULL) +
  labs(x = "Percentage-point change", y = NULL) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

## So what should you learn?

WRITE HERE. ~130 words. Choose from where demand sits, not where it looks
to be heading. Be specific.

```{r}
#| label: tbl-trends
#| tbl-cap: "Skill demand, September 2025 vs September 2026"

trend |>
  filter(n_2025 + n_2026 >= 20) |>
  transmute(Skill = skill,
            `2025 (%)`    = round(100 * share_2025, 1),
            `2026 (%)`    = round(100 * share_2026, 1),
            `Change (pp)` = round(change_pp, 1),
            p             = round(p_value, 3)) |>
  kable()
```

## Takeaway

WRITE HERE. ~60 words. One sentence a reader could repeat.

---

**Code and data:** [`posts/web-scraping/`](https://github.com/yipin-grena/aeds6400-blog/tree/main/posts/web-scraping)
